---
title: "Powerpoint - penguins"
author: "Chelsea Dack"
date: "24/11/2020"
output:
  bookdown::html_document2: default
  bookdown::word_document2: default
  bookdown::pdf_document2: default
---

<!-- set default code chunk options -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.retina = 3)
# echo = FALSE means code will not be included in output
# warning = FALSE & message = FALSE means R warnings & messages not included
# fig.retina = 3 improves figure appearance in HTML documents
```

<!-- insert code chunk for package loading -->
```{r packages}
library(tidyverse)
library(palmerpenguins)
library(GGally)
library(caret)
```

# Background
The Palmer penguins data contains size measurements for three penguin species observed on three islands in the Palmer Archipelago, Antarctica.
Introduction to palmerpenguins - https://allisonhorst.github.io/palmerpenguins/articles/intro.html

# Methods
Variables found here - https://allisonhorst.github.io/palmerpenguins/articles/intro.html
Tidy data & remove NAs

```{r tidy}
# clean variable names
penguin <- penguins_raw %>%
  janitor::clean_names()
# do str(penguins_raw) if want to see original names

# view variables we will use to see if any NAs
penguin %>% 
  select(body_mass_g,
         ends_with("_mm")) %>% 
  summary()

# filter out rows with missing values (then check any NAs again)
penguin <- penguin %>% 
  filter(!is.na(body_mass_g))

# split species into common_name & scientific_name
penguin <- penguin %>% 
  extract(species, 
          c("common_name", "scientific_name"),
          "([a-zA-Z]+\\s[a-zA-Z]+)\\s\\(([a-zA-Z]+\\s[a-zA-Z]+)\\)")
```

# Results

Explore data:

```{r explore}
# select variables of interest & pipe into ggpairs()
#outputs a matrix of pairwise plots of the variables
penguin %>%
  select(common_name, 
         sex, 
         island,
         body_mass_g,
         ends_with("_mm")) %>%
  ggpairs(aes(color = common_name))
```

## PCA of data
- 68.844% of the variation is captured in the first PC; 88.157% in the first 2 together.
```{r pca}
# select the variables & pipe into prcomp()
pca <- penguin %>% 
  select(body_mass_g,
         ends_with("_mm")) %>%
  prcomp(scale. = TRUE)
# scaling prevents variables with biggest values dominating analysis

# see the variance accounted for by each component
summary(pca)

# See the importance (loadings) of each variable in each component
pca$rotation
# PC1 = linear combination of the original variable
```

Our species separate a little better on the first two PC than any pairwise comparison.
Coloured observations by species but PCA is an unsupervised method because PCA itself takes no account of the species in creating the PCs

```{r pca plot}
# To plot, we might want to use the scores on each of the new axes and colour them by species. The scores are in a variable called $x

# Extract the scores into a dataframe with the species names
pca_labelled <- data.frame(pca$x, common_name = penguin$common_name)
# then to do a scatterplot
pca_labelled %>% 
  ggplot(aes(x = PC1, y = PC2, color = common_name)) +
  geom_point()

```

## LDA of data

```{r lda}
# Select the four variables and pipe into the MASS::lda() function which does the LDA
lda <- penguin %>% 
  select(body_mass_g,
         ends_with("_mm")) %>%
  MASS::lda(grouping = penguin$common_name)

# View the importance of each variable in each discriminant
lda$scaling

# Select the variables from Penguin that are in the model and predict the species from the lda model object
plda <- penguin %>% 
  select(body_mass_g,
         ends_with("_mm")) %>%
  predict(object = lda)
```

Caret provides us with a useful function to examine the confusion matrix. A confusion matrix is a table that tells us about the performance of a classification model. The table gives the number of:
- correct predictions: the species predicted matches the observed species
- incorrect predictions: the species predicted does not match the observed species

The confusionMatrix() function also outputs:
- Accuracy - No. correct predictions / No. of observations
- 95% CI - 95 percent confidence interval on the accuracy (using binom.test())
- No Information Rate - No. observations in the largest class / Number of observations
- P-Value [Acc > NIR] - Is the model significantly better than than you could do by always predicting the most common class (again using binom.test()).

```{r matrix}
# examine confusion matrix
confusionMatrix(plda$class, factor(penguin$common_name))
```

The model had an accuracy of 98.8304%. There were 151 Adelie Penguins of which 1 were predicted incorrectly; 68 Chinstrap Penguins of which 3 were predicted incorrectly; and 123 Gentoo Penguins of which 0 were predicted incorrectly.

```{r lda plot}
# Extract the scores into a dataframe with the species names
lda_labelled <- data.frame(plda$x,
                              common_name = penguin$common_name)

# Create a scatter plot
lda_labelled %>% 
  ggplot(aes(x = LD1, y = LD2, color = common_name)) +
  geom_point()
```

The separation between species is stronger in the LDA than in the PCA.

## LDA with training

Can use the caret function createDataPartition() to split the dataset in to training and testing sets, to avoid overfitting. It returns a proportion of row numbers randomly sampled from the dataframe.

```{r sets}
# Create a vector of row numbers that will be used to split the dataset in to training and testing sets
ids <- createDataPartition(y = penguin$common_name,
                           p = 0.75,
                           list = FALSE)
# p is the proportion of rows to sample.
# list = FALSE gives a vector of numbers rather than a one item list.

# use those row numbers to select the rows from penguin to create the training and test datasets. We use the dplyr function slice which works like the filter function but filters rows on their index rather than a match to a condition
train <- penguin %>% slice(ids)
test <- penguin %>% slice(-ids)
```

```{r lda2}
# Perform the LDA on the training data
lda <- train %>%
  select(body_mass_g,
         ends_with("_mm")) %>%
  MASS::lda(grouping = train$common_name)

# predict classes of the test data based on lda model
plda <- test %>%
  select(body_mass_g,
         ends_with("_mm")) %>%
  predict(object = lda)
```

```{r matrix2}
# Examining the confusion matrix
confusionMatrix(plda$class,factor(test$common_name))

```

