---
title: "workshop"
author: "Chelsea Dack"
date: "25/11/2020"
output:
  bookdown::word_document2: default
  bookdown::html_document2: default
  bookdown::pdf_document2: default
---

<!-- set default code chunk options -->
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      warning = FALSE,
                      message = FALSE,
                      fig.retina = 3)
# echo = FALSE means code will not be included in output
# warning = FALSE & message = FALSE means R warnings & messages not included
# fig.retina = 3 improves figure appearance in HTML documents
```

<!-- insert code chunk for package loading -->
```{r packages}
library(tidyverse)
library(Rtsne)
library(bookdown)
```

# Introduction
Dataset contains measurements of seed kernels from different varieties of wheat: Kama (1), Rosa (2) and Canadian (3). High quality visualization of the internal kernel structure was detected using a soft X-ray technique and 7 measurements were taken which are in this order in the file: Area, Perimeter, Compactness, Length of kernel, Width of kernel, Asymmetry coefficient, Length of kernel groove.

Investigate whether these variables might be useful to classifiy the species.

# Methods
```{r import}
# import data (note the file does not include column names)
file <- "data-raw/seeds_dataset.txt"
```

```{r tidy}
# create column names
cols <- c("area", 
          "perimeter",
          "compactness",
          "kernal_length",
          "kernel_width",
          "asymmetry_coef",
          "groove_length",
          "species")
seeds <- read_table2(file, col_names = cols)

# The species is coded as 1, 2, and 3 and it would be useful to recode to the species names:
seeds$species <- recode(seeds$species,
                        `1` = "Kama",
                        `2` = "Rosa",
                        `3` = "Canadian")
```

# Results
```{r ggpairs}
# If have 4 - 15 variables, plotting them pairwise against each other gives nice overview. 
# This can be achieved with the `ggpairs()` from the `GGally` package.
seeds %>% 
  GGally::ggpairs(aes(color = species)) 
```

Principal Components Analysis (PCA) is a method to investigate whether you have groups or patterns in a dataset. It is 'data reduction' or 'dimension reduction' method and creates a set of 'components' (axes) which are linear combinations of the original variables. PCA is useful when relatively few components are needed to capture most of the variation in the data.

```{r pca}
# When the values in one variable are much bigger than in others we usually scale all the variables (mean of zero and a unit variance) before undertaking PCA to avoid the variable with the biggest values dominating the analysis.
pca <- seeds %>% 
  select(-species) %>% 
  prcomp(scale. = TRUE)

# To see the variance accounted for by each component
summary(pca)
# nearly 99% variation is captured in the first three components.

# To see the importance (loading) of each variable in each component
pca$rotation
# first component seems to be an axis of size, second of shape, third of density
```

```{r pca plot}
# To plot, we might want to use the scores on each of the new axes and colour them by species. The scores are in a variable called $x
# For convenience, put these in one 'tidy' dataframe
pca_labelled <- data.frame(pca$x, species = seeds$species)

# a then to do a scatterplot
ggplot(pca_labelled, aes(x = PC1, y = PC2, color = species)) +
  geom_point()
```

t-SNE is a non-parametric, non-linear method that prioritises placing similar observations near each other. It is a probabilistic method and is computational intensive.

```{r tsne}
tsne <- seeds %>% 
  select(-species) %>%
  Rtsne(perplexity = 20,
        check_duplicates = FALSE)

dat <- data.frame(tsne$Y,  species = seeds$species)

dat %>% ggplot(aes(x = X1, y = X2, colour = species)) +
  geom_point()
```
